R Notebook

### 2023-11-15 R_code/paper_analysis.Rmd
new changes

```{r}
prefix = "/Users/noa/Workspace/ML_results"

fasttree_all_models_performance = read_tsv(paste(prefix,"/fasttree/all_models_performance.tsv", sep = ""))
fasttree_final_models_performance = read_tsv(paste(prefix,"/fasttree/final_model_performance.tsv", sep = ""))
fasttree_group_performance = read_tsv(paste(prefix,"/fasttree/groups_performance.tsv", sep = ""))
fasttree_vi =  read_tsv(paste(prefix,"/fasttree/model_standard_vi.tsv", sep = ""))
fasttree_enriched_test = read_tsv(paste(prefix,"/fasttree/test_model_standard_enriched.tsv", sep = ""))
#####
iqtree_all_models_performance = read_tsv(paste(prefix,"/iqtree/all_models_performance.tsv", sep = ""))
iqtree_final_models_performance = read_tsv(paste(prefix,"/iqtree/final_model_performance.tsv", sep = ""))
iqtree_group_performance = read_tsv(paste(prefix,"/iqtree/groups_performance.tsv", sep = ""))
iqtree_vi =  read_tsv(paste(prefix,"/iqtree/model_standard_vi.tsv", sep = ""))
iqtree_enriched_test = read_tsv(paste(prefix,"/iqtree/test_model_standard_enriched.tsv", sep = ""))
#####
raxml_all_models_performance = read_tsv(paste(prefix,"/raxml/all_models_performance.tsv", sep = ""))
raxml_final_models_performance = read_tsv(paste(prefix,"/raxml/final_model_performance.tsv", sep = ""))
raxml_group_performance = read_tsv(paste(prefix,"/raxml/groups_performance.tsv", sep = ""))
raxml_vi =  read_tsv(paste(prefix,"/raxml/model_standard_vi.tsv", sep = ""))
raxml_enriched_test = read_tsv(paste(prefix,"/raxml/test_model_standard_enriched.tsv", sep = ""))

@@

Choose specific program
```{r}
program = 'iqtree'
if (program=='fasttree') {
    all_models_performance = fasttree_all_models_performance
    final_models_performance = fasttree_final_models_performance
    group_performance = fasttree_group_performance
    vi = fasttree_vi
    enriched_test = fasttree_enriched_test
    validation_data = fasttree_validation_data

} else if (program=='iqtree') {
    all_models_performance = iqtree_all_models_performance
    final_models_performance = iqtree_final_models_performance
    group_performance = iqtree_group_performance
    vi = fasttree_vi
    enriched_test = iqtree_enriched_test
    validation_data = iqtree_validation_data
} else if (program=='raxml') {
    all_models_performance = raxml_all_models_performance
    final_models_performance = raxml_final_models_performance
    group_performance = raxml_group_performance
    vi = fasttree_vi
    enriched_test = raxml_enriched_test
    validation_data = raxml_validation_data
}
```

General performance
```{r}
final_models_performance  %>% filter (dataset=='test', metric_type=='all_data') %>% dplyr::select (analysis_type,name,balanced_accuracy_score, mcc, AUC)
```

Variable importance 
```{r}
vi %>% arrange (-`Gini-importance`)
```

@@

@@

error_vs_size<- all_models_performance %>% filter (metric_type=='all_data', dataset =='test')
error_vs_size_plt = error_vs_size %>% mutate(n_MSAs = sample_frac) %>% ggplot(aes(x=n_MSAs, y = AUC))+geom_line()+ expand_limits(y = 0.5)+xlab("Number of MSAs")+ theme(text = element_text(size = 13))+ylab("AUC")+geom_point()

@@

```
Group performance

@@

Pypythia_error<-test_group_performance %>% filter (grouping_col_name =='msa_difficulty_group') %>% ggplot(aes(y=grouping_col, x=Balanced_Accuracy))+geom_col()+ expand_limits(y = c(0.5,1))+ylab('MSA\ndifficulty')+ xlab('Balanced\naccuracy')+theme(text = element_text(size = 11))

n_seq_error<-test_group_performance %>% filter (grouping_col_name =='n_seq_group')  %>% ggplot(aes(y=grouping_col, x=Balanced_Accuracy))+geom_col()+ expand_limits(y = c(0.5,1))+ylab('Number\nof sequences')+xlab('Balanced\naccuracy')+theme(text = element_text(size = 11))

#>%  mutate(grouping_col = factor(grouping_col, levels = c('(99.999, 219.0]','(219.0, 546.0]','(546.0, 1519.0]','(1519.0, 16115.0]')))

n_loci_error<-test_group_performance %>% filter (grouping_col_name =='feature_msa_n_loci') %>% ggplot(aes(y=grouping_col, x=Balanced_Accuracy))+geom_col()+ expand_limits(y = c(0.5,1))+ylab('Number\nof positions')+xlab('Balanced\naccuracy')+theme(text = element_text(size = 11))

@@

>>>>

```{r}
prefix = "/Users/noa/Workspace/bootstrap_results/remote_results/"

fasttree_all_models_performance = read_tsv(paste(prefix,"ML_results/fasttree/all_models_performance.tsv", sep = ""))
fasttree_final_models_performance = read_tsv(paste(prefix,"ML_results/fasttree/final_model_performance.tsv", sep = ""))
fasttree_group_performance = read_tsv(paste(prefix,"ML_results/fasttree/groups_performance.tsv", sep = ""))
fasttree_vi =  read_tsv(paste(prefix,"ML_results/fasttree/final_model/full_model/model_standard_vi.tsv", sep = ""))
fasttree_enriched_test = read_tsv(paste(prefix,"ML_results/fasttree/final_model/full_model/test_model_standard_enriched.tsv", sep = ""))
fasttree_validation = read_tsv(paste(prefix,"validation_data/simulations_df_fasttree.tsv", sep = ""))
#####
iqtree_all_models_performance = read_tsv(paste(prefix,"ML_results/iqtree/all_models_performance.tsv", sep = ""))
iqtree_final_models_performance = read_tsv(paste(prefix,"ML_results/iqtree/final_model_performance.tsv", sep = ""))
iqtree_group_performance = read_tsv(paste(prefix,"ML_results/iqtree/groups_performance.tsv", sep = ""))
iqtree_vi =  read_tsv(paste(prefix,"ML_results/iqtree/final_model/full_model/model_standard_vi.tsv", sep = ""))
iqtree_enriched_test = read_tsv(paste(prefix,"ML_results/iqtree/final_model/full_model/test_model_standard_enriched.tsv", sep = ""))
iqtree_validation = read_tsv(paste(prefix,"validation_data/simulations_df_iqtree.tsv", sep = ""))
#####
raxml_all_models_performance = read_tsv(paste(prefix,"ML_results/raxml/all_models_performance.tsv", sep = ""))
raxml_final_models_performance = read_tsv(paste(prefix,"ML_results/raxml/final_model_performance.tsv", sep = ""))
raxml_group_performance = read_tsv(paste(prefix,"ML_results/raxml/groups_performance.tsv", sep = ""))
raxml_vi =  read_tsv(paste(prefix,"ML_results/raxml/final_model/full_model/model_standard_vi.tsv", sep = ""))
raxml_enriched_test = read_tsv(paste(prefix,"ML_results/raxml/final_model/full_model/test_model_standard_enriched.tsv", sep = ""))
raxml_validation = read_tsv(paste(prefix,"validation_data/simulations_df_raxml.tsv", sep = ""))

@@

Choose specific program
```{r}
program = 'fasttree'
if (program=='fasttree') {
    all_models_performance = fasttree_all_models_performance
    final_models_performance = fasttree_final_models_performance %>% filter(analysis_type!='feature_standard_fasttree_boot_support')
    group_performance = fasttree_group_performance
    vi = fasttree_vi
    enriched_test = fasttree_enriched_test
    #validation_data = fasttree_validation_data

} else if (program=='iqtree') {
    all_models_performance = iqtree_all_models_performance
    final_models_performance = iqtree_final_models_performance
    group_performance = iqtree_group_performance
    vi = iqtree_vi
    enriched_test = iqtree_enriched_test
    #validation_data = iqtree_validation_data
} else if (program=='raxml') {
    all_models_performance = raxml_all_models_performance
    final_models_performance = raxml_final_models_performance
    group_performance = raxml_group_performance
    vi = raxml_vi
    enriched_test = raxml_enriched_test
    #validation_data = raxml_validation
}
```

General performance
```{r}

out_path <-paste(prefix,"ML_results/",program,"/performance_text",".tsv", sep = "")

paper_performance<-final_models_performance  %>% filter (metric_type=='all_data') %>% dplyr::select (dataset,analysis_type,name,mcc,balanced_accuracy_score,precision, recall, AUC,average_precision)  %>% arrange(dataset) %>% mutate (description = case_when(analysis_type == "full_standard" & name =="model_standard" ~ "all_features", 
                                analysis_type == "fast_standard" & name =="model_standard" ~ "All features except NNI features",
                                analysis_type == "bootstrap_support" & name =="raw_only_boot" ~ "Raw bootstrap support",
                                analysis_type == "bootstrap_support" & name =="only_boot" ~ "Calibrated bootstrap support",
                                analysis_type == "bootstrap_support" & name =="inc_boot" ~ "All features+ bootstrap support",
                                    analysis_type == "feature_aLRT_iqtree_support" & name =="raw_only_boot" ~ "Raw aLRT support",
                                analysis_type == "feature_aLRT_iqtree_support" & name =="only_boot" ~ "Calibrated aLRT support",
                                analysis_type == "feature_aLRT_iqtree_support" & name =="inc_boot" ~ "All features+ aLRT support",
                                                                    analysis_type == "feature_aBayes_iqtree_support" & name =="raw_only_boot" ~ "Raw aBayes support",
                                analysis_type == "feature_aBayes_iqtree_support" & name =="only_boot" ~ "Calibrated aBayes support",
                                analysis_type == "feature_aBayes_iqtree_support" & name =="inc_boot" ~ "All features+ aBayes support",
                                )) %>% dplyr::select (dataset, description, mcc, balanced_accuracy_score,precision, recall)  %>% arrange(dataset)

paper_performance

Variable importance 
```{r}
out_path <-paste(prefix,"ML_results/",program,"/vi_text",".tsv", sep = "")

vi_data<- vi %>% arrange (`...1`) %>% mutate (feature_name = case_when(
  `...1` == "feature_min_ll_diff"~ "Minimum log-likelihood difference between an NNI neighbor near the bipartition and current tree",
  `...1` =="feature_max_ll_diff"~ "Maximum log-likelihood difference between an NNI neighbor near the split and current tree",

  `...1` =='feature_mean_parsimony_trees_binary'~"Fraction of parsimoy trees in which the bipartition exists",
  `...1` =='feature_mean_parsimony_trees_neighbors_binary'~"Fraction of parsimony trees in which the neighbouring bipartitions exist",

    `...1`=='feature_mean_all_ML_boot_raxml_neighbors_binary'~'Fraction of final ML trees in which the bipartition exists',

   `...1` =='feature_mean_all_ML_boot_raxml_binary'~"Fraction of parsimony trees in which the neighbouring bipartitions exist",

`...1` =='feature_mean_parsimony_trees'~"Mean transfer distance of the bipartition across parsimoy trees",
`...1` =='feature_mean_parsimony_trees_neighbors'~"Mean transfer distance of the neighbouring bipartitions across parsimoy trees",

`...1` =='feature_mean_all_ML_boot_raxml'~'Mean transfer distance of the bipartition across final ML trees',
`...1` =='feature_mean_all_ML_boot_raxml_neighbors'~'Mean transfer distance of the neighbouring bipartitions across final ML trees',

`...1` =='feature_partition_branch'~'Branch length at the parititon',
`...1` =='feature_partition_branch_vs_mean'~'Branch length at the parititon divided by total tree divergence',

`...1` =='feature_partition_divergence'~'Total divergence in the smaller subtree defined by the bipartition',
`...1` =='feature_divergence_ratio'~'Total divergence in the smaller subtree defined by the bipartition divided by total tree divergence',
`...1` =='feature_partition_size'~'Number of leaves in the smaller subtree defined by the bipartition',
`...1` =='feature_partition_size_ratio'~'Fraction of leaves in the smaller subtree defined by the bipartition',

`...1` =='feature_min_neighbor_brlen'~'Minimal branch length among the neighbouring branches',
`...1` =='feature_mean_neighbor_brlen'~'Mean branch length among the neighbouring branches',
`...1` =='feature_msa_constant_sites_pct'~'Fraction of constant sites among the MSA sites',
`...1` =='feature_msa_n_loci'~'Number of positions in the MSA',
`...1` =='feature_msa_n_unique_sites'~'Number of unique positions in the MSA',
`...1` =='feature_msa_n_seq'~'Number of sequences in the MSA',
`...1` =='feature_msa_pypythia_msa_difficulty'~'MSA difficulty',

)) %>% dplyr::select (`...1`,feature_name,`Gini-importance`, mcc ) %>% arrange(feature_name,-`Gini-importance`)
vi_data
write_tsv(vi_data,out_path)

@@

```{r}
all_models_performance
```

@@

error_vs_size<- all_models_performance %>% filter (metric_type=='all_data', dataset =='test', analysis_type=='full_standard')
error_vs_size_plt = error_vs_size %>% mutate(n_MSAs = sample_frac*2007) %>% ggplot(aes(x=n_MSAs, y = mcc))+geom_line()+ expand_limits(y = 0)+xlab("Number of MSAs")+ theme(text = element_text(size = 13))+ylab("MCC")+geom_point()

@@

```
```{r}
test_group_performance %>% filter (grouping_col_name =='feature_msa_n_loci') 
```

Group performance

@@

Pypythia_error<-test_group_performance %>% filter (grouping_col_name =='msa_difficulty_group') %>%  mutate(grouping_col = factor(grouping_col, levels = c('(0.00328, 0.303]','(0.303, 0.538]','(0.538, 0.571]','	(0.571, 0.865]'))) %>%
ggplot(aes(y=grouping_col, x=Balanced_Accuracy))+geom_col()+ expand_limits(y = c(0.5,1))+ylab('MSA\ndifficulty')+ xlab('Balanced\naccuracy')+theme(text = element_text(size = 11))

n_seq_error<-test_group_performance %>% filter (grouping_col_name =='n_seq_group')   %>%  mutate(grouping_col = factor(grouping_col, levels = c('(4.999, 69.0]','(69.0, 129.0]','129.0, 236.0]','	(236.0, 827.0]]'))) %>% ggplot(aes(y=grouping_col, x=Balanced_Accuracy))+geom_col()+ expand_limits(y = c(0.5,1))+ylab('Number\nof sequences')+xlab('Balanced\naccuracy')+theme(text = element_text(size = 11))

n_loci_error<-test_group_performance %>% filter (grouping_col_name =='feature_msa_n_loci') %>%  mutate(grouping_col = factor(grouping_col, levels = c('(4.999, 69.0]','(69.0, 129.0]','129.0, 236.0]','	(236.0, 827.0]]')))  %>% ggplot(aes(y=grouping_col, x=Balanced_Accuracy))+geom_col()+ expand_limits(y = c(0.5,1))+ylab('Number\nof positions')+xlab('Balanced\naccuracy')+theme(text = element_text(size = 11))

@@

```{r}
cor.test( enriched_test$feature_msa_pypythia_msa_difficulty,  enriched_test$prob_predictions)
cor.test( enriched_test$feature_msa_n_seq, enriched_test$prob_predictions)
cor.test( enriched_test$feature_msa_n_loci, enriched_test$prob_predictions)
```

```{r}
running_time_data<- enriched_test%>% dplyr::select (boot_run_time,extraction_running_time,nni_ll_eval_running_time) %>% mutate (diff1 = boot_run_time/extraction_running_time, diff2 = boot_run_time/(extraction_running_time+nni_ll_eval_running_time
)) 
summary(running_time_data %>% pull (diff1))
summary(running_time_data %>% pull (diff2))
summary(running_time_data %>% pull (extraction_running_time))
summary(running_time_data %>% pull (nni_ll_eval_running_time))
#enriched_test %>% dplyr::select (boot_run_time, extraction_runni
```

```{r}

running_time_data<- enriched_test%>% dplyr::select (program_boot_run_time,extraction_running_time,nni_ll_eval_running_time) %>% mutate (diff1 = program_boot_run_time/extraction_running_time, diff2 = program_boot_run_time/(extraction_running_time+nni_ll_eval_running_time
)) 
summary(running_time_data %>% pull (diff1))
summary(running_time_data %>% pull (diff2))
summary(running_time_data %>% pull (extraction_running_time))
summary(running_time_data %>% pull (nni_ll_eval_running_time))
summary(running_time_data %>% pull (program_boot_run_time))
#enriched_test %>% dplyr::select (boot_run_time, extraction_running_time, nni_ll_eval_running_time)
```
```{r}
summary(validation_data$feature_n_unique_seq)
summary(enriched_test$feature_msa_n_seq)
```
